Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-12 intro.tex
a few edits to intro.tex

As a corollary of the undecidability of the halting problem
\cite{halting-problem}, however, no single forecasting schema is ideal
for all noise-free deterministic signals\cite{weigend-book}---let
alone all real-world time series data sets.  This leads naturally to
an important and challenging question: given a complicated real-valued
time series, does there exist a forecast model that can utilize the
information (if any) that is being transmitted from past to future by
the underlying generating process?  A first step in answering this
question is to reliably quantify where on the complexity spectrum a
given time series falls; a second step is to determine how complexity
and predictability are related.  These are the goals of this paper.

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to
predict.  An important practical corollary to this is a strategy for
assessing appropriateness of forecast methods.  If the forecast
produced by a particular method is poor but the time series contains a
significant amount of predictive structure, that method is inadequate
to the task and one should seek another method.  We focus on
real-valued, scalar, time-series data that appear complicated: i.e.,
not periodic, linear, constant, etc.  We make no assumptions about the
properties of the generating process: whether it is linear, nonlinear,
deterministic, stochastic, stationary, non-stationary, etc.

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation\cite{crutchfield2003}.  Our approach
exploits this decomposition in order to assess how much predictive
structure is present in a signal---i.e., where it falls on the
complexity spectrum mentioned above.  We define \emph{complexity} as a
particular approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.
That is, we view a random-walk time series (which exhibits high
entropy) as purely complex, whereas a periodic signal, which exhibits
low entropy, is on the low end of the complexity spectrum.  We argue
that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy through
ordinal analysis---is an effective way to assess the complexity of a
given time series.  Permutation entropy is ideal for our purposes
because it is known to converge to the true entropy value. Other
existing techniques either require specific knowledge of the
generating process or produce biased values of the
entropy~\cite{bollt2001}.

To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a number of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is
  quantifiable by permutation entropy and correlated with prediction
  accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema---and in the choice of which forecasting scheme is appropriate.
\end{enumerate}
The forecast methods used here were chosen to span the space of
standard prediction strategies, but they do not cover that space
exhaustively.  Our goal here is an empirical assessment of the
relationship between predictability and complexity, not formal results
about a ``best'' predictor for a given time series.  From a
practitioner's standpoint, it would be useful to know {\sl a priori}
if a trace contains enough predictive structure to make it worth
spending the time and effort searching for a good forecast method.  It
would also be useful to know if a given method is inadequate---that
is, if one could do better and should keep searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  Computers are among the most complex
engineered artifacts in current use.  Modern microprocessor chips
contain multiple processing units and multi-layer memories, for
instance, and they use complicated hardware/software strategies to
move data and threads of computation across those resources.  As a
result, the processor and memory loads during the execution of even a
very simple program can evolve in a very complicated fashion.
Figure~\ref{fig:ipc} shows a performance trace of a three-line loop
that repeatedly initializes the upper triangle of a matrix in
column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic \cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the program can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting its \emph{performance}, which does not violate the
  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

@@

The rest of the paper is organized as follows:
Section~\ref{sec:related} discusses previous results on generating
partitions, local modeling, and error distribution analysis, and
situates this work in that context. Section~\ref{sec:methods} covers
the experimental setup and methods used to collect each time
series. Section~\ref{sec:model} describes the prediction models we
implemented.  In Section~\ref{sec:meaComplex}, we review weighted
permutation entropy.  In Section~\ref{sec:results}, we estimate the
complexity of each time series and compare that complexity to the
accuracy of predictions produced by the methods of
Section~\ref{sec:model} on that time series.  In
Section~\ref{sec:conc}, we discuss these results and their
implications and consider future areas of research.

>>>>

As a corollary of the undecidability of the halting problem
\cite{halting-problem}, however, no single forecasting schema is ideal
for all noise-free deterministic signals\cite{weigend-book}---let
alone all real-world time-series data sets.  This leads naturally to
an important and challenging question: given a complicated real-valued
time series, does there exist a forecast model that can utilize the
information (if any) that is being transmitted from past to future by
the underlying generating process?  A first step in answering this
question is to reliably quantify where on the complexity spectrum a
given time series falls; a second step is to determine how complexity
and predictability are related.  These are the goals of this paper.

In particular, we wish to determine---without knowing anything about
the generating process---whether a time series is too complex to
predict.  An important practical corollary to this is a strategy for
assessing appropriateness of forecast methods.  If the forecast
produced by a particular method is poor but the time series contains a
significant amount of predictive structure, one can reasonably conclude that that method is inadequate
to the task and that one should seek another method.

The information in an observation can be partitioned into two pieces:
redundancy and entropy generation\cite{crutchfield2003}.  Our approach
exploits this decomposition in order to assess how much predictive
structure is present in a signal---i.e., where it falls on the
complexity spectrum mentioned above.  We define \emph{complexity} as a
particular approximation of Kolmogorov-Sinai entropy\cite{KS-entropy}.
That is, we view a random-walk time series (which exhibits high
entropy) as purely complex, whereas a periodic signal, which exhibits
low entropy, is on the low end of the complexity spectrum.  We argue
that an extension of \emph{permutation entropy}
\cite{bandt2002per}---a method for approximating the entropy through
ordinal analysis---is an effective way to assess the complexity of a
given time series.  Permutation entropy is ideal for our purposes
because it works with real-valued data and is known to converge to the true entropy value. Other
existing techniques either require specific knowledge of the
generating process or produce biased values of the
entropy~\cite{bollt2001}.

We focus on real-valued, scalar, time-series data that appear
complicated: i.e., not periodic, linear, constant, etc.  We make no
assumptions about the properties of the generating process: whether it
is linear, nonlinear, deterministic, stochastic, stationary,
non-stationary, etc.
To explore the relationship between complexity, predictive structure,
and predictability, we generate forecasts of a variety of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is {\sl (i)}
  quantifiable by permutation entropy and {\sl (ii)} correlated with prediction
  accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema---and in the choice of which one is appropriate for a given 
time series.
\end{enumerate}
The forecast methods used in this study were chosen to span the space of
standard prediction strategies, but they do not, of course, cover that space
exhaustively.  Our goal here is an empirical assessment of the
relationship between predictability and complexity, not formal results
about a ``best'' predictor for a given time series.  From a
practitioner's standpoint, it would be useful to know {\sl a priori}
if a trace contains enough predictive structure to make it worth
spending the time and effort searching for a good forecast method.  It
would also be useful to know if a given method is inadequate---that
is, if one could do better and should keep searching.

@@

For the purposes of this study, we require a broad array of
time-series datasets that cover the full complexity spectrum.  Rather
than use synthetic data, we study sensor data from computer
performance experiments.  While this is not a common laboratory
experiment, it is a highly appropriate choice here.  Computers are
among the most complex engineered artifacts in current use.  Modern
microprocessor chips contain multiple processing units and multi-layer
memories, for instance, and they use complicated hardware/software
strategies to move data and threads of computation across those
resources.  As a result, the processor and memory loads during the
execution of even a very simple program can evolve in a very
complicated fashion.  Figure~\ref{fig:ipc} shows a performance trace
of a three-line loop that repeatedly initializes the upper triangle of
a matrix in column-major order.
%
 \begin{figure}[htbp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a three-line C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:ipc}
  \end{figure}
%
Although this time series appears to be periodic, it is actually
chaotic \cite{mytkowicz09}, which places fundamental limits on
predictability.  Changes in the program can radically change this
behavior; indeed, many performance traces exhibit interesting regime
changes as a program moves through the different phases of its
operation.  The resulting traces span the whole range of the
complexity spectrum, from completely predictable to completely
unstructured, making this an ideal testbed for this
study\footnote{Predicting the \emph{state} of a computer, of course,
  would amount to solving the halting problem.  What we are doing here
  is predicting computer \emph{performance}, which does not violate the
  Rice-Shapiro theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

@@

The rest of the paper is organized as follows.
Section~\ref{sec:related} discusses previous results on generating
partitions, local modeling, and error distribution analysis, and
situates this work in that context. Section~\ref{sec:methods} covers
the experimental setup and methods used to collect each time
series. Section~\ref{sec:model} describes the prediction models used
in this study.  In Section~\ref{sec:meaComplex}, we review the
permutation entropy technique.  In Section~\ref{sec:results}, we estimate the
complexity of each time series and compare that complexity to the
accuracy of predictions produced by the methods of
Section~\ref{sec:model}, operating on that time series.  In
Section~\ref{sec:conc}, we discuss these results and their
implications and consider future areas of research.
